The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-16 talk/vmil2012/paper.tex
(cfbolz, bivab) lstinline settings

\lstset{
  basicstyle=\ttfamily\footnotesize,
  language=Python,
  keywordstyle=\bfseries,
  stringstyle=\color{blue},
  commentstyle=\color{commentgray}\textit,
  fancyvrb=true,
  showstringspaces=false,
  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
  numberstyle = \tiny,
  numbersep = -20pt,
}

@@

@@

%___________________________________________________________________________
\todo{mention somewhere that it is to be expected that most guards do not fail}
\todo{better formatting for lstinline}
\section{Introduction}

\todo{the introduction needs some work}
\cfbolz{the first two two paragraphs talk about deoptimization, then it
switches to guards. I would say we should only talk about guards in the
beginning}
Tracing just-in-time (JIT) compilers record and compile linear control flow paths of operations executed by an interpreter
inserting operations called guards at points of possible divergence.
In this paper we describe and analyze how guards work, explaining what concepts
are used in the intermediate and low-level representation of the JIT
instructions and how these are implemented.

Our aim is to help understand the constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points, meaning the recorded
and compiled path has to be left returning control to the interpreter. They need
enough associated information to enable rebuilding the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, to explain how they are implemented in the different levels of this particular
tracing JIT and to clarify the rationale behind the design decisions based on the
numbers provided here.

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section \ref{sec:Resume Data}). During the recording phase guards are
inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after
optimizing the traces generated for the different benchmarks used in this paper.
This makes guards one of the most common types of operations. As this paper will show, many of these guards
fail rarely or not all during execution.

There are several aspects to be taken into account w.r.t. guards, the first of them is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the construction of a tracing just-in-time compiler.

@@

The set of central concepts upon which this work is based are described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend. In this context the frontend is concerned with recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the machine specific JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

The RPython language and the PyPy project were started in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension.\footnote{\url{http://pypy.org}} PyPy is now a fully compatible
alternative implementation of the Python language, that is on average about 5 times faster than the reference implementation. The
implementation takes advantage of RPython's language features
such as the provided tracing just-in-time compiler described below.

RPython, the language and the toolset originally created to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations. Besides
the Python interpreter there are several language implementation at different
levels of completeness, e.g. for Smalltalk, Prolog, JavaScript and R.

RPython is constructed from two components:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language
is a statically typed object oriented high level language. The language provides
several features such as automatic memory management
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the program to a low level representations suited to be compiled
and run on one of the different supported target platforms/architectures such
as C, .NET and Java. During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as (if needed) a garbage collector
and based on hints provided by the author a just-in-time compiler.

@@

After the recorded trace has been optimized it is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can merged, further reducing the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how the \lstinline{int_eq} operation
followed by a \lstinline{guard_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

>>>>

\lstdefinelanguage{none}{
      keywords={},
}
\lstset{
  basicstyle=\ttfamily\footnotesize,
  language=none,
  keywordstyle=\bfseries,
  stringstyle=\color{blue},
  commentstyle=\color{commentgray}\textit,
  fancyvrb=true,
  showstringspaces=false,
  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
  numberstyle = \tiny,
  numbersep = -20pt,
}

@@

\let\oldlstinline=\lstinline
\renewcommand\lstinline[1]{\oldlstinline[basicstyle=\ttfamily]{#1}}

@@

%___________________________________________________________________________
\section{Introduction}

Tracing just-in-time (JIT) compilers record and compile commonly executed
linear control flow paths consisting of operations executed by an interpreter.
At points of possible divergence from the traced path operations called guards
are inserted. Furthermore, type guards are inserted to specialize the trace
based on the types observed during tracing. In this paper we describe and
analyze how guards work and explain the concepts used in the intermediate and
low-level representation of the JIT instructions and how these are implemented.
This is done in the context of the RPython language and the PyPy project, which
provides a tracing JIT compiler geared at dynamic language optimization.

Our aim is to help understand the constraints when implementing guards
and to describe the concrete techniques used in the various layers of RPython's
tracing JIT. All design decisions will be motivated by concrete numbers for the
frequency and the overhead related to guards.

It is important to handle guards well, because they are very common operations
in the traces produced by tracing JITs. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after optimizing
the traces generated for the different benchmarks used in this paper. An
additional property is that guard failure rates are very uneven. The majority
of guards never fail at all, whereas those that do usually fail extremely
often.

Besides being common, guards have various costs attached to them.
Guards have a runtime cost, they take time to execute. Therefore it is
important to make the on-trace execution of guards as efficient as possible. On
the other hand, guards are possible deoptimization points. The recorded and
compiled path has to be left if a guard fails, returning control to the
interpreter. Therefore guards need enough associated information to enable
rebuilding the interpreter state. The memory overhead of this information
should be kept low. These constraints and trade-offs are what make the design
and optimization of guards an important and non-trivial aspect of the construction
of a tracing just-in-time compiler.

@@

The set of central concepts upon which this work is based are described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend. In this context the frontend is concerned with recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section~\ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the machine specific JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

The RPython language and the PyPy project~\cite{rigo_pypys_2006} were started
in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension.\footnote{\url{http://pypy.org}} PyPy is now a fully compatible
alternative interpreter for the Python language.
Using RPython's tracing JIT compiler it is on average about 5 times faster than
CPython, the reference implementation.
PyPy is an interpreter written in RPython and takes advantage of the language
features provided by RPython
such as the provided tracing just-in-time compiler described below.

RPython, the language and the toolset originally created to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations. Besides
the Python interpreter there are several experimental language implementation at different
levels of completeness, e.g. for Prolog~\cite{bolz_towards_2010}, Smalltalk~\cite{bolz_towards_2010}, JavaScript and R.

different levels of completeness.

RPython can mean one of two things:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language
is a statically typed object-oriented high level language. The language provides
several features such as automatic memory management
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the interpreter into a C program.\footnote{
    RPython can also be used to translate programs to CLR and Java
    bytecode~\cite{ancona_rpython:_2007}, but this feature is somewhat
    experimental.
}
During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as a garbage collector and a tracing JIT compiler.
The process of inserting a tracing JIT is not fully automatic but is guided by
hints from the interpreter author.

@@

After the recorded trace has been optimized it is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can merged, further reducing the overhead of the guard.
Figure~\ref{fig:trace-compiled} shows how the \lstinline{int_eq} operation
followed by a \lstinline{guard_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.
